Attrition in Massive Open Online Courses: Inequalities, Dropout Reasons, and Psychological Factors

### 2014-10-30 paper_v1.Rnw
started general discussion, need to eat
Fixes and changes

Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc,guthrie2013moocs}. Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}. Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs}, motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested. Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized. Kizilcec and Schneider \citeyear{kizilcec2015motivation} proposed that MOOCs have bridged two different world: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'', which presupposes instructor-defined goals that students strive to achieve \cite{tyack1994grammar}. Viewing MOOC participation as bridging these two worlds undoubtedly adds a layer of complexity to interpretations of attrition. The present work extends the empirical base of research on attrition to inform the theoretical discourse and pave the path for novel practices.

This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 20,879 online learners in 18 courses. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals. We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes. Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. In Study 2, to develop a deeper understanding of MOOC attrition, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback. Their open response answers about what challenges they faced while taking the MOOC were iteratively coded. We investigated the role of volitional control for learners who report having not enough time, and examine goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.

@@

The particular MOOC under observation was an undergraduate level course on an advanced topic in computer science. It was offered in 2014 through Coursera. There were  20,048 enrolled learners; 10,510 watched at least one video, and [N] attempted more than one assignment. The following subsections describe how disengaging learners were identified, which measures were included in the feedback survey, and how open responses were iteratively coded.

@@

The objective was to build an `early warning' prediction model that could flag learners who are likely to drop out in the future. We defined the following three criteria by which to evaluate the performance of the model.

@@

The training set was composed of learners' interaction data (predictors) and dropout states (outcome) from 20 MOOCs. The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses. Five hundred learners from each MOOC were included in the training set ($train$), and 500 other learners from the same MOOCs were selected for the first test set ($test_1$). A second test set of similar size was composed from 20 other MOOCs ($test_2$). Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see \cite{halawa2014dropout}, for additional details on the prediction model). The model was fit by logistic regression. Table \ref{tab:drop_pred_rocs} provides information on the prediction performance, measured by the area under the ROC curve (AUC).

@@

>>>>

Shortly after the first wave of courses had finished, extensive media coverage led to MOOCs becoming associated with high attrition rates \cite{lewin2013after,parr2013mooc,guthrie2013moocs}. Early MOOC research cautioned against dichotomizing learners into successes and failures based on course completion \cite{kizilcec2013deconstructing,rivard2013measuring}. Instead, more nuanced categorizations based on learner behavior \cite{kizilcec2013deconstructing,clow2013moocs}, motivations \cite{kizilcec2015motivation}, or intentions \cite{wilkowski2014student} have been suggested. Ultimately, perspectives on persistence and attrition in MOOCs depend on how MOOCs have been conceptualized. Kizilcec and Schneider \citeyear{kizilcec2015motivation} proposed that MOOCs have bridged two different worlds: one is governed by the user-centric norms of online media, where everyone is encouraged to be as active as they wish; the other adheres to the ``grammar of schooling'', which presupposes instructor-defined goals that students strive to achieve \cite{tyack1994grammar}. Viewing MOOC participation as bridging these two worlds undoubtedly adds a layer of complexity to interpretations of attrition. The present work extends the empirical base of research on attrition to inform the theoretical discourse and pave the path for novel practices.

This paper presents a systematic investigation of attrition in MOOCs, based on self-report and behavioral data collected from 66,076 online learners in 18 courses. A better understanding of why learners disengage from MOOCs allows for earlier recognition of impending attrition, which enables targeted interventions to help learners achieve their learning goals. We begin by briefly reviewing the large literature on attrition in educational environments with a focus on important developments in understanding its causes. Building on this foundation of prior work, Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. In Study 2, to develop a deeper understanding of MOOC attrition, we invited around 6,500 learners who were predicted to have disengaged from a MOOC to provide feedback. Their open response answers about what challenges they faced while taking the MOOC were iteratively coded. We investigated the role of volitional control for learners who report having not enough time, and examine goal striving, social belonging, and growth mindset in relation to being successful in a MOOC.

@@

The particular MOOC under observation was an undergraduate level course on an advanced topic in computer science. It was offered in 2014 through Coursera. There were  20,048 enrolled learners; 10,510 watched at least one video, and 5,912 attempted at least one assignment. The following subsections describe how disengaging learners were identified, which measures were included in the feedback survey, and how open responses were iteratively coded.

@@

The objective was to build an `early warning' prediction model that could flag learners who are likely to have dropped out or to drop out in the future. In order for our predictions to be useful, we need to observe the following three features of the dropout prediction problem.

@@

The training set was composed of learners' interaction data (predictors) and dropout states (outcome) from 20 MOOCs. The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses. Five hundred learners from each MOOC were included in the training set ($train$), and 500 other learners from the same MOOCs were selected for the first test set ($test_1$). A second test set of similar size was composed from 20 other MOOCs ($test_2$). Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see \cite{halawa2014dropout}, for additional details on the prediction model).

To constrain prediction lag, we stopped feeding the model with a learner's activity data once they have been absent for 14 consecutive days. During training, if the learner had dropped out but the model had not detected her within 14 days of her last appearnce, the learner is counted as a miss (false negative). This construction causes the training algorithm to choose from among features that only operate within the prediction lag constraints we define.

The model was fit by logistic regression. Table \ref{tab:drop_pred_rocs} provides information on the prediction performance, measured by the area under the ROC curve (AUC).

@@

The almost identical results obtained for $test_1$ and $test_2$ imply good transferability of the model.
